Fast Einsum Implementation Using Parallelization, Vectorization, and More\\\large Algorithm Engineering 2025 Project Paper

### 2024-12-28 Exam Assignments/paper/paper.tex
verbatim -> lstlisting

\documentclass[sigconf]{acmart}
\usepackage{booktabs}

@@

% The five-finger pattern \cite{macgilchrist2014}:
\begin{enumerate}
% \item \textbf{Topic and background:} What topic does the paper deal with? What is the point of departure for your research? Why are you studying this now?
\item \textbf{Topic and background:} This paper concerns tensor operations, specifically those implemented in the Python library numpy with its \texttt{numpy.einsum} method. Tenor operations are very common in todays data driven world, especially with the rise of \em Deep Learning\em . Hence, we should ensure that those important methods are implemented efficiently, as \em time \em and \em space complexity \em are a big concern when operating on increasingly bigger data.

@@

In order to not always hardcode these operations, \em einsum \em can be employed by passing in the two operand tensors, together with a format string, which defines how every dimension will be treated. For example \texttt{ik,kj->ij} specifies a normal matrix-matrix-multiplication. Here is how these \em column identifiers \em of \texttt{lhs\_string} in the eisnum expression \texttt{<lhs\_string>,<rhs\_string>-><target\_string>} are interpreted:

\begin{table}[h!]
  \centering
  \caption{Column Identifiers of Einsum Expression}
  \label{tab:column_identifiers}
  \begin{tabular}{lccc}
    \toprule
    \textbf{Category} & \textbf{rhs\_string} & \textbf{target\_string} \\
    \midrule
    Batch              & Yes    & Yes    \\
    Kept left          & No     & Yes    \\
    Contracted         & Yes    & No     \\
    Summed left        & No     & No     \\
    \bottomrule
  \end{tabular}
\end{table}

You have to read the table in the following manner. Take a column identifier of \texttt{lhs\_string}. Then you check whether it is present in \texttt{rhs\_string} and \texttt{target\_string}, and based on these results you assign it to one of the four specified categories, where "Summed left" describes a dimension to reduce.

@@

~\\
Next, we have to ensure that the tuple \texttt{(std::string, Tensor<T>, Tensor<T>)} is a valid einsum expression. To this end, I implemented \texttt{is\_valid\_einsum\_expression} function found in \texttt{include/compute\_einsum.h}. It checks the following five criteria:

@@

In case the tuple passed all these checks, the function won't throw any error, and hence we can now proceed with the function \texttt{compute\_einsum} knowing we have a valid einsum expression.

@@

\subsubsection{Transposing and Reshaping Operand Tensors}
Now that we have categorized all the dimensions, we bring both tensors into the canonical form

\[ \text{lhs\_tensor}[\vec{b}, \vec{k_l}, \vec{c}, \vec{s_l}] \ , \ \text{rhs\_tensor}[\vec{b}, \vec{k_r}, \vec{c}, \vec{s_r}] \quad . \]

One very important detail is that the order of the \em batch \em and \em contracted \em dimensions after transposing have the same ordering in the operand tensors. To illustrate this, say we have the einsum expression \texttt{einsum("ab,ba->", lhs\_matrix, rhs\_matrix)}. Naively, there is no need to transpose both matrices, as both of them are in the canonical form (every dimension is a contraction dimension). However, the orderings of these contracted dimensions do not match, hence we do need to transpose a matrix, either the left or the right hand side.

~\\
To this end, the algorithm reorders the indices-lists for the right hand side such that the resulting order of column identifiers matches the left hand side. For this purpose, we employ the function \texttt{column\_identifiers\_like}.

~\\
Furthermore, it is worth mentioning that there are different ways to transpose tensors. One way is the "lazy" way, where we just alter meta attributes of the tensor, in particular we alter the \em strides\em , which specify by "how much" we have to multiply an index of a given dimension in order to get the physical index to access the element in the underlying data array. This lazy way of doing it is very fast, as there is no need to move large amount of data. However, there is one big problem: If a tensor has non-canonical strides, we cannot reshape it. This is very bad, because we actually do want to reshape the tensors afterwards. Hence, we need to use the other way of transposing tensors: In he "eager" approach we basically move the data such that it is aligned with the new ordering of the dimensions. This is obviously much more expensive time and space complexity wise, but there is at least one benefit: The data is well aligned for the next operation (after reshaping): The following reduction operation will benefit from a stride access pattern of 1.

@@

We start by reducing the fourth dimension of both operand tensors. This is fairly efficient as we have efficient access patterns as described earlier. After that, the two tensors will have the following form:

\[ \text{lhs\_tensor}[\vec{b}, \vec{k_l}, \vec{c}] \ , \ \text{rhs\_tensor}[\vec{b}, \vec{k_r}, \vec{c}] \quad . \]

It is worth mentioning that it might be possible to combine all of these operations so far (transposing, reshaping, reducing fourth dimension) into one big one. This will at least save some memory, as now we have three tensors for each operand tensor in memory: The original one which we have to leave untouched, the transposed one, as well as the reduced one. I still decided to implement it this way, as this approach is more modular. Furthermore, I initially implemented a transpose method myself, but I realized that it was the bottleneck of my implementation. Hence, I decided to use an external library for transposing, such that the performance of my implementation is not limited by a poor implemented transposing function.

@@

\subsubsection{Batch Matrix Multiplication}
Now we take care of contracting the appropriate dimensions by performing a batch matrix multiplication. Note, though, that the contraction dimensions are the last one in each operand tensor (the third dimension), and hence we actually perform a \em batch matrix matmul transpose \em operation, i.e. we treat the rhs operand batch matrices as being transposed. I chose this implementation, as this way we have a stride of one when multiplying individual matrices of a batch both in the lhs and rhs matrix, with respect to the inner most loop. Consequently, this ensures excellent spatial locality and hence we make perfect use of the cache. The result of this operation will be the three dimensiol tensor \texttt{bmm\_result}, with dimension sizes

\[ \text{bmm\_result} = [b, k_l, k_r] \quad . \]

As hinted, we might store the result of this operation in previously allocated memory, all provided that the memory region is big enough. Otherwise we do have to allocate new memory. The question whether our available memory is sufficient for the new data depends on on the dimension sizes of the operand tensors and their categorized dimensions: We have

\begin{align*}
|\text{lhs\_tensor}| &= b \cdot k_l \cdot c \cdot s_l \\
|\text{rhs\_tensor}| &= b \cdot k_r \cdot c \cdot s_r \\
|\text{bmm\_result}| &= b \cdot k_l \cdot k_r \quad ,
\end{align*}

and hence the memory will be sufficiently big enough iff

\[ c \cdot s_l \geq k_r \]

in case that the lhs tensor is the bigger one of the two (similar if rhs tensor is bigger).

@@

To illustrate this, say we have a matrix a 5x10 matrix $M$, and we want to transpose it. When accessing $M[i][j]$ we have to calculate the physical index of the row-major order stored matrix. Hence:

\[ \text{physical\_index}_M(i, j) = 10 \cdot i + 1 \cdot j \quad . \]

When we now transpose this matrix, we can alter this translation function instead of actually moving the data. In this example:

\[ \text{physical\_index}_{M^T}(i, j) = 1 \cdot i + 10 \cdot j \quad . \]

This means that instead of moving data, we transpose the coefficients of this function. These coefficients are stored as \em strides \em in the metadata of a tensor. NumPy does so as well, and hence we employ this by doing this final transposition the lazy manner by setting these NumPy metadata attributes.

This implementation choice is not "cheating", as NumPy's einsum implementation does the same lazy transposition. Hence, it is only fair to do so myself, as this way we get more meaningful comparisons. To prove my point, run the following lines of Python:

\begin{verbatim}
import numpy as np
matrix = np.array([[1,2],[3,4]])
print(matrix.strides)
transposed = np.einsum("ab,c->ba", matrix, np.array([1]))
print(transposed.strides)
\end{verbatim}

You will likely see that the strides got altered simply.

@@

\begin{verbatim}
#pragma omp parallel for
for (size_t i = 0; i < target_size; i++) {
    T sum = 0;
    size_t start_idx = i * last_dimension;
    size_t end_idx = start_idx + last_dimension;

    for (size_t k = start_idx; k < end_idx; k++) {
        sum += this->data[k];
    }

    target_data[i] = sum;
}
\end{verbatim}

Clearly, this code is very efficient. There are two minor problems though: First, there is a potential risk of overflow when summing up the values. Second, we parallelize the outer most loop, which is a wise choice, but in very rare cases when \texttt{target\_size} equals one, like when reducing a one-dimensional vector, there won't be any parallelization.

@@

% TODO: Explain vectorized versions
Despite that, I actually tried to optimize this method by using explicit vectorization in specialized template methods. Interestingly, they don't seem to have any performance benefits compared to the generic implementation, which is probably due to the fact that modern compilers will vectorize the generic implementation as well. The code is pretty much self explanatory, we do employ an expensive \texttt{loadu} instead of a \texttt{load} vector instruction, as the data won't be aligned generally.

@@

\begin{verbatim}
#pragma omp parallel for collapse(3)
for (size_t b = 0; b < batch_size; ++b) {
    for (size_t i = 0; i < rows; ++i) {
        for (size_t j = 0; j < cols; ++j) {
            const T* lhs_row = lhs.data + b * rows * inner_dim + i * inner_dim;
            const T* rhs_col = rhs.data + b * cols * inner_dim + j * inner_dim;
            T* result_row = result_data + b * rows * cols + i * cols;

            T sum = T(); // Initialize sum as zero for type T
            for (size_t k = 0; k < inner_dim; ++k) {
                sum += lhs_row[k] * rhs_col[k];
            }
            result_row[j] = sum;
        }
    }
}
\end{verbatim}

We face the \textbf{exact} problems we faced at the reduction method, and my responses are the same, we just keep the implementation simple for the sake of demonstration. This function could be manually vectorized as well, but the compiler should do a decent job at optimizing this function, and hence we shouldn't face a big performance penalty.

@@

There are several Improvements one could make, but the most obvious ones are:
\begin{itemize}
  \item Augmenting supported data types. Therefore a new implementation of the transposition method is needed.
  \item More support for parameters, maybe also tensors of different data types, different operations, etc.
  \item More optimized main method, more heuristics, etc.
  \item More optimized sub methods, improving parallelization, vectorization by checking for aligned data, etc.
\end{itemize}

@@

Table~\ref{tab:results} shows the running times of the resolution step of the five best placed teams.

\begin{table}[htbp]
  \caption{Comparison of the F-measure and the running times of the resolution step of the five best placed teams. The input data for the resolution step consisted of 29{,}787 in JSON formatted e-commerce websites. Measurements were taken on a
laptop running Ubuntu 19.04 with 16 GB of RAM and two Intel Core i5-4310U CPUs. The underlying SSD was a 500\,GB 860 EVO mSATA. We cleared the page cache, dentries, and inodes before each run to avoid reading the input data from RAM instead of the SSD.}
  \label{tab:results}
\resizebox{\columnwidth}{!}{
  \begin{tabular}{lcrr}
    \toprule
    Team& Language & F-measure & Running time (s)\\
    \midrule
	PictureMe (\textbf{this paper}) &C++& 0.99 & \textbf{0.61}\\
    DBGroup@UniMoRe &Python& 0.99 & 10.65\\
    DBGroup@SUSTech &C++& 0.99 & 22.13\\
    eats\_shoots\_and\_leaves &Python& 0.99 & 28.66\\
    DBTHU &Python& 0.99& 63.21\\
  \bottomrule
\end{tabular}
}
\end{table}

\section{Conclusions}
\label{sec:conclusions}

>>>>

\documentclass[sigconf]{acmart}
\usepackage{booktabs}
\usepackage{pgfplotstable}
\usepackage{fancyvrb}
\usepackage{listings}
\usepackage{etoolbox}

\lstset{
    breaklines=true       % Enable line breaking
}
% Define a custom lstlisting command that avoids page breaks
\BeforeBeginEnvironment{lstlisting}{\noindent\begin{minipage}{\linewidth}}
\AfterEndEnvironment{lstlisting}{\end{minipage}}

% TODO: write acronyms like cpu, gpu capitalized

\setlength{\emergencystretch}{3em} % Allow LaTeX to stretch lines slightly before exceeding width
\tolerance=1000                   % Allow more "badness" in line breaks before exceeding width
\hbadness=10000                   % Report warnings for badness levels higher than this value

\pretolerance=2000  % Try harder to find good breaks before a paragraph
\tolerance=3000     % Relax line-breaking rules further
\emergencystretch=5em % Allow additional stretch space
\renewcommand{\texttt}[1]{\begingroup\ttfamily\sloppy\hbadness=10000 #1\endgroup}
\linepenalty=1000 % Increase the penalty for exceeding width
\sloppy

@@

% The five-finger pattern \cite{macgilchrist2014}:
\begin{enumerate}
% \item \textbf{Topic and background:} What topic does the paper deal with? What is the point of departure for your research? Why are you studying this now?
\item \textbf{Topic and background:} This paper concerns tensor operations, specifically those implemented in the Python library numpy with its \texttt{numpy.einsum} method. Tensor operations are very common in todays data driven world, especially with the rise of \em Deep Learning\em . Hence, we should ensure that those important methods are implemented efficiently, as \em time \em and \em space complexity \em are a big concern when operating on increasingly bigger data.

@@

\noindent
In order to not always hardcode these operations, \em einsum \em can be employed by passing in the two operand tensors, together with a format string, which defines how every dimension will be treated. For example \texttt{ik,kj->ij} specifies a normal matrix-matrix-multiplication. Here is how these \em column identifiers \em of \texttt{lhs\_string} in the eisnum expression \texttt{<lhs\_string>,<rhs\_string>-><target\_string>} are interpreted:

\begin{table}[h!]
  \centering
  \caption{Column Identifiers of Einsum Expression}
  \label{tab:column_identifiers}
  \begin{tabular}{lccc}
    \toprule
    \textbf{Category} & \textbf{rhs\_string} & \textbf{target\_string} \\
    \midrule
    Batch              & Yes    & Yes    \\
    Kept left          & No     & Yes    \\
    Contracted         & Yes    & No     \\
    Summed left        & No     & No     \\
    \bottomrule
  \end{tabular}
\end{table}
\noindent
You have to read the table in the following manner. Take a column identifier of \texttt{lhs\_string}. Then you check whether it is present in \texttt{rhs\_string} and \texttt{target\_string}, and based on these results you assign it to one of the four specified categories, where "Summed left" describes a dimension to reduce.

@@

Next, we have to ensure that the tuple \texttt{(std::string, Tensor<T>, Tensor<T>)} is a valid einsum expression. To this end, I implemented \texttt{is\_valid\_einsum\_expression} function found in \texttt{include/compute\_einsum.h}. It checks the following five criteria:

@@

\noindent
In case the tuple passed all these checks, the function won't throw any error, and hence we can now proceed with the function \texttt{compute\_einsum} knowing we have a valid einsum expression.

@@

\subsubsection{Transposing and Reshaping Operand Tensors}
Now that we have categorized all the dimensions, we bring both tensors into the canonical form
\[ \text{lhs\_tensor}[\vec{b}, \vec{k_l}, \vec{c}, \vec{s_l}] \ , \ \text{rhs\_tensor}[\vec{b}, \vec{k_r}, \vec{c}, \vec{s_r}] \quad . \]
One very important detail is that the order of the \em batch \em and \em contracted \em dimensions after transposing have the same ordering in the operand tensors. To illustrate this, say we have the einsum expression \texttt{einsum("ab,ba->", lhs\_matrix, rhs\_matrix)}. Naively, there is no need to transpose both matrices, as both of them are in the canonical form (every dimension is a contraction dimension). However, the orderings of these contracted dimensions do not match, hence we do need to transpose a matrix, either the left or the right hand side.

To this end, the algorithm reorders the indices-lists for the right hand side such that the resulting order of column identifiers matches the left hand side. For this purpose, we employ the function \texttt{column\_identifiers\_like}.

Furthermore, it is worth mentioning that there are different ways to transpose tensors. One way is the "lazy" way, where we just alter meta attributes of the tensor, in particular we alter the \em strides\em , which specify by "how much" we have to multiply an index of a given dimension in order to get the physical index to access the element in the underlying data array. This lazy way of doing it is very fast, as there is no need to move large amount of data. However, there is one big problem: If a tensor has non-canonical strides, we cannot reshape it. This is very bad, because we actually do want to reshape the tensors afterwards. Hence, we need to use the other way of transposing tensors: In he "eager" approach we basically move the data such that it is aligned with the new ordering of the dimensions. This is obviously much more expensive time and space complexity wise, but there is at least one benefit: The data is well aligned for the next operation (after reshaping): The following reduction operation will benefit from a stride access pattern of 1.

@@

We start by reducing the fourth dimension of both operand tensors. This is fairly efficient as we have efficient access patterns as described earlier. After that, the two tensors will have the following form
\[ \text{lhs\_tensor}[\vec{b}, \vec{k_l}, \vec{c}] \ , \ \text{rhs\_tensor}[\vec{b}, \vec{k_r}, \vec{c}] \quad . \]
It is worth mentioning that it might be possible to combine all of these operations so far (transposing, reshaping, reducing fourth dimension) into one big one. This will at least save some memory, as now we have three tensors for each operand tensor in memory: The original one which we have to leave untouched, the transposed one, as well as the reduced one. I still decided to implement it this way, as this approach is more modular. Furthermore, I initially implemented a transpose method myself, but I realized that it was the bottleneck of my implementation. Hence, I decided to use an external library for transposing, such that the performance of my implementation is not limited by a poor implemented transposing function.

@@

\subsubsection{Batch Matrix Multiplication}
Now we take care of contracting the appropriate dimensions by performing a batch matrix multiplication. Note, though, that the contraction dimensions are the last one in each operand tensor (the third dimension), and hence we actually perform a \em batch matrix matmul transpose \em operation, i.e. we treat the rhs operand batch matrices as being transposed. I chose this implementation, as this way we have a stride of one when multiplying individual matrices of a batch both in the lhs and rhs matrix, with respect to the inner most loop. Consequently, this ensures excellent spatial locality and hence we make perfect use of the cache. The result of this operation will be the three dimensiol tensor \texttt{bmm\_result}, with dimension sizes
\[ \text{bmm\_result} = [b, k_l, k_r] \quad . \]
As hinted, we might store the result of this operation in previously allocated memory, all provided that the memory region is big enough. Otherwise we do have to allocate new memory. The question whether our available memory is sufficient for the new data depends on on the dimension sizes of the operand tensors and their categorized dimensions: We have
\begin{align*}
|\text{lhs\_tensor}| &= b \cdot k_l \cdot c \cdot s_l \\
|\text{rhs\_tensor}| &= b \cdot k_r \cdot c \cdot s_r \\
|\text{bmm\_result}| &= b \cdot k_l \cdot k_r \quad ,
\end{align*}
and hence the memory will be sufficiently big enough iff
\[ c \cdot s_l \geq k_r \]
in case that the lhs tensor is the bigger one of the two (similar if rhs tensor is bigger).

@@

To illustrate this, say we have a matrix a 5x10 matrix $M$, and we want to transpose it. When accessing $M[i][j]$ we have to calculate the physical index of the row-major order stored matrix. Hence:
\[ \text{physical\_index}_M(i, j) = 10 \cdot i + 1 \cdot j \quad . \]
When we now transpose this matrix, we can alter this translation function instead of actually moving the data. In this example:
\[ \text{physical\_index}_{M^T}(i, j) = 1 \cdot i + 10 \cdot j \quad . \]
This means that instead of moving data, we transpose the coefficients of this function. These coefficients are stored as \em strides \em in the metadata of a tensor. NumPy does so as well, and hence we employ this by doing this final transposition the lazy manner by setting these NumPy metadata attributes.

This implementation choice is not "cheating", as NumPy's einsum implementation does the same lazy transposition. Hence, it is only fair to do so myself, as this way we get more meaningful comparisons. To prove my point, run the following lines of Python:
\begin{lstlisting}[language=Python]
import numpy as np
matrix = np.array([[1,2],[3,4]])
print(matrix.strides)
transposed = np.einsum("ab,c->ba", matrix, np.array([1]))
print(transposed.strides)
\end{lstlisting}
You will likely see that the strides got altered simply.

@@

\begin{lstlisting}[language=C++]
#pragma omp parallel for simd aligned(data: 32) aligned(target_data: 32)
for (size_t i = 0; i < target_size; i++) {
    T sum = 0;
    size_t start_idx = i * last_dimension;
    size_t end_idx = start_idx + last_dimension;

    for (size_t k = start_idx; k < end_idx; k++) {
        sum += data[k];
    }

    target_data[i] = sum;
}
\end{lstlisting}
\\\\
Clearly, this code is very efficient. There are two minor problems though: First, there is a potential risk of overflow when summing up the values. Second, we parallelize the outer most loop, which is a wise choice, but in very rare cases when \texttt{target\_size} equals one, like when reducing a one-dimensional vector, there won't be any parallelization.

@@

Despite that, I actually tried to optimize this method by using explicit vectorization in specialized template methods. Interestingly, they don't seem to have any performance benefits compared to the generic implementation, which is probably due to the fact that modern compilers will vectorize the generic implementation as well. The code is pretty much self explanatory, we do employ an efficient \texttt{load} instead of a \texttt{loadu} vector instruction, as the data will be aligned because of the previous transpose operation. Here is the code for tensors of type \texttt{float}:

\begin{lstlisting}[language=C++]
#pragma omp parallel for
for (size_t i = 0; i < target_size; i++) {
    float sum = 0;
    size_t start_idx = i * last_dimension;

    // Use AVX SIMD for vectorized summation
    __m256 vec_sum = _mm256_setzero_ps(); // Initialize to zero

    for (size_t k = start_idx; k < start_idx + last_dimension; k += 8) {
        // Load 8 values at once
        __m256 vec_vals = _mm256_load_ps(&this->data[k]);
        vec_sum = _mm256_add_ps(vec_sum, vec_vals); // Add the values
    }

    // Sum the partial results
    sum += vec_sum[0] + vec_sum[1] + vec_sum[2] + vec_sum[3]
        +  vec_sum[4] + vec_sum[5] + vec_sum[6] + vec_sum[7];

    target_data[i] = sum;
}
\end{lstlisting}

@@

\begin{lstlisting}[language=C++]
#pragma omp parallel for simd aligned(result_data: 32) aligned(lhs_data: 32) aligned(rhs_data: 32) collapse(3)
for (size_t b = 0; b < batch_size; ++b) {
    for (size_t i = 0; i < rows; ++i) {
        for (size_t j = 0; j < cols; ++j) {
            const T* lhs_row = lhs_data + b * rows * inner_dim + i * inner_dim;
            const T* rhs_col = rhs_data + b * cols * inner_dim + j * inner_dim;
            T* result_row = result_data + b * rows * cols + i * cols;
            T sum = T(); // Initialize sum as zero for type T

            #pragma omp simd
            for (size_t k = 0; k < inner_dim; ++k) {
                sum += lhs_row[k] * rhs_col[k];
            }
            result_row[j] = sum;
        }
    }
}
\end{lstlisting}
\\\\
We face the \textbf{exact} problems we faced at the reduction method, and my responses are the same, we just keep the implementation simple for the sake of demonstration. This function could be manually vectorized as well, but the compiler should do a decent job at optimizing this function, and hence we shouldn't face a big performance penalty.

@@

There are several Improvements one could make, but the most obvious ones are:
\begin{itemize}
  \item Augmenting supported data types. Therefore a new implementation of the transposition method is needed.
  \item More support for parameters, maybe also tensors of different data types, different operations, etc.
  \item More optimized main method, more heuristics like only transposing when needed, etc.
  \item More optimized sub methods, improving parallelization, vectorization by checking for aligned data, etc.
  \item Utilize gpu for parallelization.
\end{itemize}

@@

Now that everything is implemented, we test our einsum implementation and compare it to NumPy's. To this end, I have written \texttt{einspeed.py}, which compares both implementations on a variety of test cases focussing on different aspects of einsum calculations.

\subsection{Correctness}
We start by assessing the correctness. Our implementation seems to operate very precisely, as for small problem sizes both implementations yield the same output. For larger problem sizes, however, there is a growing discrepancy in the entries of the result tensor. This discrepancy can grow arbitrary large for increasing tensor sizes, which is why a relative measure is more meaningful. We calculate it like this:

\begin{lstlisting}[language=Python]
maximum_relative_discrepancy = np.max(np.abs((result - correct) / correct))
\end{lstlisting}
\\\\
Note that per constraction no entry of \texttt{correct} is zero.
It turns out that this relative discrepancy stays below a value of approximately \texttt{1e-4}, which is very commendable. One might ask where this error arises, and I my answer to that will be that numerical operations of floating point numbers don't form a mathematical \em field\em . Consequently, the order of operation will have an effect on the final output, and I think that my implementation employs a different order than NumPy's.

Ultimately, however, neither implementanion is exactly correct, and my implementation isn't necessarily worse. In fact, on some test cases it was closer to a hard coded answer, but this doesn't have to mean anything, as these computations will suffer from numerical errors as well. Just to give an example, one can test this by running:

\begin{lstlisting}[language=Python]
lhs_vector = np.random.rand(100)
rhs_vector = np.random.rand(100)
print(np.einsum("i,i->", lhs_vector, rhs_vector))
print(einspeed.einsum("i,i->", lhs_vector, rhs_vector))
print(lhs_vector @ rhs_vector)
\end{lstlisting}
\\\\
In conclusion, my implementation seems to operate correctly, and hence we now focus on performance.

\subsection{Performance}
When running the test cases, is is apparent that NumPy did a really good job at implementing einsum, as it outperforms my multi-threaded implementation on many test cases. I think this is mostly due to the fact that my memory management is not the most efficient as I use a lot of memory, and NumPy probably uses a bit more sophisticated algorithms for the computations as well. Notably, there seems to be only a constant factor that separates my implementation from NumPy's, and this factor is roughly 20.

These findings emphasize that the developers behind NumPy really know what they are doing. Hence, it is no surprise that it is very challenging to implement certain methods more efficient than they have. Still, it is very interesting to deeper understand the mechanism behind some implementations, and to test out the limits of computations myself.

Very interestingly, NumPy's einsum function seems to \textbf{really} struggle at big reduction operations, which is very surprising, as they are very simple to implement. This is one of the key findings of this paper. There seems to be a time complexity issue with its implementation, as with growing operand sizes the running time becomes horrendously bad. To give evidence, we execute this einsum expression
\begin{lstlisting}[language=Python]
einsum("abc,def->", lhs_tensor, rhs_tensor)
\end{lstlisting}
at different operand scales, and compare both implementations. The results are shown in table~\ref{tab:results}.

\begin{table}[htbp]
  \caption{Comparison of einsum running times. Both NumPy's and my einspeed implementation are tested at varying problem sizes. The problem size specifies the number of entries in the two input tensors. Running times are measured in seconds, the speedup describes the fraction $\frac{numpy\_time}{einspeed\_time}$. 
  Measurements were taken on a PC running Ubuntu 24.04 with 32 GB of RAM and an AMD Ryzen 5 5600G CPU (6 Cores, 12 Threads).}
  \label{tab:results}
\resizebox{\columnwidth}{!}{
  \pgfplotstabletypeset[
    col sep=comma,
    every head row/.style={% Styling for the header row
      before row=\toprule, 
      after row=\midrule},
    every last row/.style={% Styling for the last row
      after row=\bottomrule},
  ]{src/scaling_study.csv}
}
\end{table}

\section{Conclusions}
\label{sec:conclusions}
In this paper, I implemented the einsum function from scratch, in order to evaluate the existing NumPy implementanion. Despite the fact that I certainly could have done better, I was pretty satisfied with my work at the end. The fact that my implementation can "keep up" with NumPy's contributed to this. Yet, I couldn't outperform NumPy generally, only in very specific cases. To be honest, that was predictable. Therefore, these findings emphasize the quality of of libraries like NumPy, but they also show that we should be cautious about performance when using these functions for special purposes, like big reduction operations. Based on my results I advise the reader not to use NumPy's einsum implementation for big reduction operations, as this will result in a serious bottleneck. One should hardcode this operation instead, or use a different einsum implementation. For the latter case, one might use the results of this paper as a starting point to build a better einsum implementation.
